---
title: "MSG Frame Coverage Analysis"
author: "Felix Baez-Santiago"
date: "2026-05-26"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(dplyr)
library(janitor)
library(tidyr)
library(tidyverse)
library(stringr)
library(knitr)
library(kableExtra)

msg_census_data0 = read.csv("O:/slteam/SMART/NCSES MSG/msg_addresses_wide.csv")

msg_census_data = msg_census_data0 %>%
  filter(
    pumaid %in% c("2600100", "2600802", "2601200", "2601600", "2601701", "2601703", "2602903", "2603203", "2603212")
  ) %>% 
  mutate(pumaid = as.character(pumaid))
```

## Overview
-1 how much does MSG overcount occupied households relative to 
census benchmarks
-2 is that discrepancy systematic with respect to 
county-level characteristics?

## 1. MSG vs. Census Household Counts by County

MSG household counts are compared against ACS and Decennial occupied unit counts.  A ratio above 1.0 indicates MSG overcounts relative to the benchmark.

```{r coverage-table}
msg_hh_county <- msg_census_data %>% 
  group_by(fips) %>%
  summarise(
    msg_hh_count   = n_distinct(msgid),
    msg_vacant     = n_distinct(msgid[vacantcode == "Y"]),
    msg_occupied   = n_distinct(msgid[vacantcode == "N"]),
    .groups = "drop"
  )

# ACS and Decennial counts by county (total and occupied)
acs_hh_county <- msg_census_data %>%
  distinct(fips_bg, .keep_all = TRUE) %>%
  group_by(fips) %>%
  summarise(
    acs_hu_total  = sum(hu_total_bgE,        na.rm = TRUE),
    acs_occ_total = sum(occ_occupied_bgE,    na.rm = TRUE),
    dec_hu_total  = sum(dec_occ_total_bg,    na.rm = TRUE),
    dec_occ_total = sum(dec_occ_occupied_bg, na.rm = TRUE),
    .groups = "drop"
  )

 # Join and compute ratios vs both total HU and occupied HU
county_coverage <- msg_hh_county %>%
  left_join(acs_hh_county, by = "fips") %>%
  mutate(
    # vs occupied
    msg_vs_acs_occ_ratio = msg_hh_count / acs_occ_total,
    msg_vs_dec_occ_ratio = msg_hh_count / dec_occ_total,
    # vs total housing units
    msg_vs_acs_hu_ratio  = msg_hh_count / acs_hu_total,
    msg_vs_dec_hu_ratio  = msg_hh_count / dec_hu_total
  )

tab1 <- county_coverage %>%
  select(fips, msg_hh_count, msg_vacant, msg_occupied,
         acs_hu_total, acs_occ_total,
         dec_hu_total, dec_occ_total,
         msg_vs_acs_occ_ratio, msg_vs_acs_hu_ratio,
         msg_vs_dec_occ_ratio, msg_vs_dec_hu_ratio) %>%
  mutate(across(c(msg_vs_acs_occ_ratio, msg_vs_acs_hu_ratio,
                  msg_vs_dec_occ_ratio, msg_vs_dec_hu_ratio), ~round(., 2))) %>%
  kable(col.names = c("FIPS", 
                      "MSG Total", "MSG Vacant", "MSG Occupied",
                      "ACS Total", "ACS Occ",
                      "Dec Total", "Dec Occ",
                      "MSG/ACS Occ", "MSG/ACS Tot",
                      "MSG/Dec Occ", "MSG/Dec Tot"),
        caption = "MSG vs. Census HH Counts by County (Total and Occupied)",
        format.args = list(big.mark = ",")) %>%
  kable_styling(font_size = 8, latex_options = c("scale_down", "hold_position"))
tab1
```



---

## 2. Does Overcoverage Depend on County Characteristics?

```{r chars-correlation}

county_chars <- msg_census_data %>%
  distinct(fips_bg, .keep_all = TRUE) %>%
  group_by(fips) %>%
  summarise(
    pov_rate     = sum(pov_below_bgE,     na.rm = TRUE) / sum(pov_total_bgE,    na.rm = TRUE),
    pct_hispanic = sum(race_hispanic_bgE, na.rm = TRUE) / sum(race_total_bgE,   na.rm = TRUE),
    pct_rented   = sum(tenure_rented_bgE, na.rm = TRUE) / sum(tenure_total_bgE, na.rm = TRUE),
    median_inc   = weighted.mean(median_hhincome_bgE, hu_total_bgE, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as.data.frame()

county_coverage_chars <- county_coverage %>%
  left_join(county_chars, by = "fips") %>%
  as.data.frame()

cor_matrix <- county_coverage_chars %>%
  select(msg_vs_acs_occ_ratio, msg_vs_acs_hu_ratio,
         msg_vs_dec_occ_ratio, msg_vs_dec_hu_ratio,
         pov_rate, pct_hispanic, pct_rented, median_inc) %>%
  cor(use = "complete.obs") %>%
  round(3)

short_names <- c("MSG/ACS Occ", "MSG/ACS Total", 
                 "MSG/Dec Occ", "MSG/Dec Total",
                 "Poverty Rate", "% Hispanic", 
                 "% Rented", "Median Income")

rownames(cor_matrix) <- short_names
colnames(cor_matrix) <- short_names

tab2 <- cor_matrix %>%
  kable(caption = "Correlations between MSG Overcoverage Ratios and County Characteristics") %>%
  kable_styling(font_size = 9, latex_options = c("scale_down", "hold_position"))

tab2
```


# ACS PUMS Household-Level Indicators
```{r}
# read in person level pums file
library(googledrive)
library(readr)

plink <- "https://drive.google.com/file/d/1V24rheRA0pjVl5JNZEVSFv8nRF6wbmp0/view?usp=drive_link"

# Extract file ID from link
file_id <- sub(".*?/d/([^/]+)/.*", "\\1", plink)

# Get drive file object
drv_file <- drive_get(as_id(file_id))

# Download to a temporary local file
tmp <- tempfile(fileext = ".csv")   # change extension if needed
drive_download(drv_file, path = tmp, overwrite = TRUE)

# Read into R
psam_p26 <- read_csv(tmp) %>% 
  filter(
    PUMA %in% c( "00100", "00802", "01200", "01600", "01701", "01703", "02903", "03203", "03212")
  ) 

# read in household level pums file
hlink ="https://drive.google.com/file/d/16Qbt4R0xLeEB_m3YlQGJ7qIq550MOO3X/view?usp=sharing"

# Extract file ID from link
file_id <- sub(".*?/d/([^/]+)/.*", "\\1", hlink)

# Get drive file object
drv_file <- drive_get(as_id(file_id))

# Download to a temporary local file
tmp2 <- tempfile(fileext = ".csv")   # change extension if needed
drive_download(drv_file, path = tmp2, overwrite = TRUE)

# Read into R
psam_h26 <- read_csv(tmp2) %>% 
  filter(
    PUMA %in% c( "00100", "00802", "01200", "01600", "01701", "01703", "02903", "03203", "03212")
  ) 
```

```{r}
# restrict household file to actual housing units (drop group quarters)
psam_h26_hu <- psam_h26 %>%
  filter(TYPEHUGQ == 1)

# Marital status indicator (married couple household vs. not)
psam_h26_hu <- psam_h26_hu %>%
  mutate(
    married_couple_hh = case_when(
      HHT == 1        ~ 1L,
      HHT %in% 2:7     ~ 0L,
      TRUE             ~ NA_integer_
    )
  )

psam_h26_hu %>% tabyl(married_couple_hh, show_na = TRUE) %>% adorn_pct_formatting(digits = 2)
```

```{r}
#  Count of household members aged 65+
hh_65plus <- psam_p26 %>%
  group_by(SERIALNO) %>%
  summarise(n_65plus = sum(AGEP >= 65, na.rm = TRUE), .groups = "drop")

psam_h26_hu <- psam_h26_hu %>%
  left_join(hh_65plus, by = "SERIALNO") %>%
  mutate(n_65plus = replace_na(n_65plus, 0))

summary(psam_h26_hu$n_65plus)
```


```{r}
# household income allocation flag
psam_h26_hu <- psam_h26_hu %>%
  mutate(
    hhincome_imputed = case_when(
      FHINCP == 1 ~ 1L,
      FHINCP == 0 ~ 0L,
      TRUE        ~ NA_integer_
    )
  )

psam_h26_hu %>% tabyl(hhincome_imputed, show_na = TRUE) %>% adorn_pct_formatting(digits = 2)
```


```{r}
# Limited English proficiency variable
psam_h26_hu <- psam_h26_hu %>%
  mutate(
    lep_household = case_when(
      LNGI == 2 ~ 1L,  
      LNGI == 1 ~ 0L, 
      TRUE      ~ NA_integer_
    )
  )

psam_h26_hu %>% tabyl(lep_household, show_na = TRUE) %>% adorn_pct_formatting(digits = 2)
```


```{r}
puma_pums_summary <- psam_h26_hu %>%
  group_by(PUMA) %>%
  summarise(
    n_households       = n(),
    pct_married        = round(mean(married_couple_hh, na.rm = TRUE) * 100, 2),
    avg_65plus         = round(mean(n_65plus, na.rm = TRUE), 2),
    pct_income_imputed = round(mean(hhincome_imputed, na.rm = TRUE) * 100, 2),
    pct_lep            = round(mean(lep_household, na.rm = TRUE) * 100, 2),
    .groups = "drop"
  )
tab_pums_summary <- puma_pums_summary %>%
  kable(col.names = c("PUMA", "N Households", "% Married",
                      "Avg. 65+", "% Income Imputed", "% LEP"),
        caption = "PUMA-Level Summary: PUMS-Derived Household Indicators",
        format.args = list(big.mark = ",")) %>%
  kable_styling(font_size = 9, latex_options = c("scale_down", "hold_position"))
tab_pums_summary
```

## 3. PUMA-County Structure
```{r}
library(readxl)

# Joy's county-PUMA relationship classification
mapping <- as.data.frame(read_xlsx("C:/Users/felbasa/Downloads/NCSES MSG/mi_puma_county_indicator.xlsx"))

county_puma_counts <- mapping %>%
  group_by(county_fips) %>%
  summarise(n_pumas = n_distinct(puma_code), .groups = "drop")

puma_county_counts <- mapping %>%
  group_by(puma_code) %>%
  summarise(n_counties = n_distinct(county_fips), .groups = "drop")

county_relationship <- mapping %>%
  left_join(county_puma_counts, by = "county_fips") %>%
  left_join(puma_county_counts, by = "puma_code") %>%
  group_by(county_fips) %>%
  summarise(
    n_pumas        = first(n_pumas),
    max_n_counties = max(n_counties),
    .groups = "drop"
  ) %>%
  mutate(
    relationship = case_when(
      n_pumas == 1 & max_n_counties == 1 ~ "1:1",
      n_pumas  > 1 & max_n_counties == 1 ~ "1:n",
      n_pumas == 1 & max_n_counties  > 1 ~ "n:1",
      n_pumas  > 1 & max_n_counties  > 1 ~ "n:m"
    ),
    fips = as.integer(county_fips)
  )

# --- Step 1: PUMA-level MSG + ACS comparison (eligible counties only) ---
# NOTE: kept UNROUNDED here on purpose — rounding happens once, at display time (Step 4)

puma_bg_raw <- msg_census_data %>%
  distinct(fips_bg, .keep_all = TRUE) %>%
  left_join(county_relationship %>% select(fips, relationship), by = "fips") %>%
  filter(relationship %in% c("1:1", "n:1"))

puma_msg_counts <- msg_census_data %>%
  left_join(county_relationship %>% select(fips, relationship), by = "fips") %>%
  filter(relationship %in% c("1:1", "n:1")) %>%
  group_by(pumaid) %>%
  summarise(
    msg_hh_count = n_distinct(msgid),
    msg_occupied = n_distinct(msgid[vacantcode == "N"]),
    .groups = "drop"
  )

puma_acs_chars <- puma_bg_raw %>%
  group_by(pumaid) %>%
  summarise(
    acs_occ_total = sum(occ_occupied_bgE, na.rm = TRUE),
    acs_hu_total  = sum(hu_total_bgE,     na.rm = TRUE),
    pov_rate      = sum(pov_below_bgE,     na.rm = TRUE) / sum(pov_total_bgE,    na.rm = TRUE),
    pct_hispanic  = sum(race_hispanic_bgE, na.rm = TRUE) / sum(race_total_bgE,   na.rm = TRUE),
    pct_rented    = sum(tenure_rented_bgE, na.rm = TRUE) / sum(tenure_total_bgE, na.rm = TRUE),
    median_inc    = weighted.mean(median_hhincome_bgE, hu_total_bgE, na.rm = TRUE),
    .groups = "drop"
  )

puma_comparison_table <- puma_msg_counts %>%
  left_join(puma_acs_chars, by = "pumaid") %>%
  mutate(
    msg_vs_acs_occ_ratio = msg_occupied / acs_occ_total,
    pumaid = as.character(pumaid)   # fix type for downstream joins
  )

# --- Step 2: within-PUMA poverty variance significance test ---

puma_county_map <- msg_census_data %>%
  distinct(pumaid, fips) %>%
  arrange(pumaid, fips) %>%
  left_join(county_relationship %>% select(fips, relationship), by = "fips") %>%
  filter(relationship %in% c("1:1", "n:1"))

puma_pov_vals <- puma_county_map %>%
  left_join(county_coverage_chars %>% select(fips, pov_rate), by = "fips") %>%
  group_by(pumaid) %>%
  summarise(pov_vals = list(pov_rate), .groups = "drop")

var_chisq_pval <- function(data) {
  if (length(data) < 2 || all(is.na(data))) return(NA_real_)
  data    <- data[!is.na(data)]
  df      <- length(data) - 1
  v       <- var(data)
  sigma0  <- 1e-6
  testchi <- df * v / (sigma0^2)
  pchisq(testchi, df = df, lower.tail = FALSE)
}

puma_sd_flag <- puma_pov_vals %>%
  mutate(
    sd_pov = sapply(pov_vals, function(x) sd(x, na.rm = TRUE)),
    pval   = sapply(pov_vals, var_chisq_pval),
    pumaid = as.character(pumaid)   # fix type for join
  ) %>%
  select(pumaid, sd_pov, pval)

puma_comparison_table <- puma_comparison_table %>%
  left_join(puma_sd_flag, by = "pumaid")

# --- Step 3: PUMS-derived indicators ---
# puma_pums_summary must already exist (built earlier from psam_h26_hu)

puma_pums_summary <- puma_pums_summary %>%
  mutate(pumaid = paste0("26", PUMA))

puma_master_table <- puma_comparison_table %>%
  left_join(
    puma_pums_summary %>% select(pumaid, n_households, pct_married, avg_65plus,
                                  pct_income_imputed, pct_lep),
    by = "pumaid"
  )

# --- Step 4: round ONCE, format for display ---

puma_master_display <- puma_master_table %>%
  mutate(
    pov_rate              = round(pov_rate, 3),
    pct_hispanic          = round(pct_hispanic, 3),
    pct_rented            = round(pct_rented, 3),
    msg_vs_acs_occ_ratio  = round(msg_vs_acs_occ_ratio, 2),
    sd_label = case_when(
      is.na(sd_pov)              ~ "—",
      !is.na(pval) & pval < 0.05 ~ paste0(round(sd_pov, 3), " *"),
      TRUE                       ~ as.character(round(sd_pov, 3))
    )
  ) %>%
  select(pumaid, msg_hh_count, msg_occupied, acs_occ_total, acs_hu_total,
         msg_vs_acs_occ_ratio, pov_rate, sd_label, pct_hispanic, pct_rented,
         median_inc, pct_married, avg_65plus, pct_income_imputed, pct_lep)

tab_puma_master <- puma_master_display %>%
  kable(col.names = c("PUMA ID", "MSG Total", "MSG Occupied",
                      "ACS Occ", "ACS Total", "MSG/ACS Occ Ratio",
                      "Poverty Rate", "SD Pov. Rate", "% Hispanic", "% Rented",
                      "Median Income", "% Married", "Avg. 65+", "% Income Imputed", "% LEP"),
        caption = "PUMA-Level Summary: MSG Coverage, ACS Characteristics, and PUMS Indicators.
                   * SD significantly greater than zero (chi-square, p < 0.05).
                   Only 1:1 and n:1 counties included to avoid double counting.",
        format.args = list(big.mark = ",")) %>%
  kable_styling(font_size = 8, latex_options = c("scale_down", "hold_position"))

tab_puma_master
```